---
title: "R Notebook"
output: html_notebook
---

# PHS Scottish Stroke Statistics

This notebook contains analysis steps of stroke incidence and mortality 
data across Scotland.

* Raw datasets are from PHS open data accessible via: https://www.opendata.nhs.scot/dataset/scottish-stroke-statistics
* The raw datasets have been cleaned and saved within the folder "../clean_data/"
  the cleaned datafiles will be used and analysed here.

```{r}
# load in libraries
library(tidyverse)
library(skimr)
```

### Contents

For the analysis I focus on:

1. Stroke activity
  - demographics (age, sex)
  - distribution across Scotland
2. Stroke mortality
  - demographics (age, sex)
  - distribution across Scotland

### 1. Load in cleaned datafiles

```{r}
#stroke activity per healthboard
activity_hb <- read_csv("../clean_data/stroke_activity_healthboard.csv")

#stroke activity per council area
activity_ca <- read_csv("../clean_data/stroke_activity_council.csv")

# stroke mortality per healthboard
mortality_hb <- read_csv("../clean_data/stroke_mortality_healthboard.csv")

# stroke mortality per council area
mortality_ca <- read_csv("../clean_data/stroke_mortality_council.csv")

# additional datafile 'table2_mortality_stroke_2023_01_24.csv' from
# https://publichealthscotland.scot/publications/scottish-stroke-statistics/scottish-stroke-statistics-year-ending-31-march-2022/
# showing mortality rates per SIMD quantile
mortality_simd <- read_csv("../clean_data/cd_mortality_by_deprivation.csv")
```

### 2. General exploration data contents
Initially we focus on Health boards as data contents are similar except for area.

```{r}
# general exploration activity

head(activity_hb)
view(activity_hb)

activity_hb %>% 
  filter(age_group == "All") %>% 
  filter(sex == "All")

activity_hb %>% 
  distinct(financial_year)
# data is spanning from 2012/2013 until 2021/22 = 10 years

activity_hb %>% 
  distinct(admission_type)
# 4 admission types: elective, emergency, transfer and all

activity_hb %>% 
  distinct(age_group)
# 6 age groups: 0-44, 45-64, 65-74, 75+, all, under 75

activity_hb %>% 
  distinct(diagnosis)
# 4 diagnosis types: Cerebrovascular disease, stroke, Subarachnoid haemorrhage, TIAs and related symptoms

activity_hb %>% 
  distinct(sex)
# 3 sex types: Males, Females, All

activity_hb %>% 
  distinct(hbr)
# 14 healthboards and single code for whole of Scotland

```


```{r}
# general exploration mortality
# most categories for character data are similar for activity and mortality datasets, except:

mortality_hb %>% 
  distinct(diagnosis)
# 3 diagnosis types: Cerebrovascular disease, stroke, Subarachnoid haemorrhage

```

__crude rate and easr__
Two columns in all PHS open data files are of interest for data visualisation:

*crude rate:*
The crude rate is the population-standardised rate of the 'event' (here: incidences or moratlity rate)
Crude rate is the number of events per 100.000 people

*EASR:*
is the The European Age-Sex Standardised Rate per 100,000 population.
The numbers of deaths per 100 000 population are influenced by the age distribution of the population.
use is made of the European Standard Population (ESP) to age and sex standardise.

### Visualisations - are organised per data file / topic

3a = stroke activity per HB
3b = stroke mortality per HB
3c = mortality per simd (Scottish Index Multiple Deprivation)
4a = stroke activity per HB - demographics
4b = stroke activity per HB - demographics

#### 3A. General insights Stroke activity per HB

__What is the number of total stroke incidences in Scotland per year from 2012 to now?__
```{r}
activity_hb %>% 
  select(financial_year, hbr, age_group, sex, admission_type, diagnosis, number_of_discharges, crude_rate, easr) %>% 
  filter(age_group == "All",
         admission_type == "All",
         sex == "All",
         hbr == "S92000003") %>% 
  group_by(financial_year) %>% 
  summarise(total_stroke_incidences = sum(number_of_discharges)) %>% 
  ggplot() +
  aes(x = financial_year, y = total_stroke_incidences) +
  geom_line(aes(group = 1), colour = "indianred", size = 2) +
  theme_light() +
  labs(
    x = "\nYear", 
    y = "Total incidences",
    title = "Total Cerebrovascular Disease Incidences per Year\n"
  ) +
  scale_y_continuous(limits = c(0, 70000))

```
Data/graph shows the total number of stroke incidences in Scotland over the years 
2012/13 - 2021/22. We see an initial increase until 2016 after incidence rates are 
more stable.

However this gives the absolute number of incidences, which are not standardised for 
age and sex composition differences in the population across years.
The population size changes and you want to standardise for this.

__age-sex standardised stroke incidences in Scotland per year__
```{r}
activity_hb %>% 
  select(financial_year, hbr, age_group, sex, admission_type, diagnosis, number_of_discharges, crude_rate, easr) %>% 
  filter(age_group == "All",
         admission_type == "All",
         sex == "All",
         hbr == "S92000003") %>% 
  group_by(financial_year) %>% 
  summarise(total_incidences_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = financial_year, y = total_incidences_easr) +
  geom_line(aes(group = 1), colour = "indianred", size = 2) +
  theme_light() +
  labs(
    x = "\nYear", 
    y = "age-sex adjusted incidence rate per 100,000 population",
    title = "Cerebrovascular Disease Standardised Incidences per Year",
    subtitle = "age and sex adjusted incidence rates per year"
  ) +
  scale_y_continuous(limits = c(0, 1300))
  
```


__What is the number of total stroke incidences split per diagnosis type?__

```{r}
plot_diagnosis_incidences <- activity_hb %>% 
  select(financial_year, hbr, age_group, sex, admission_type, diagnosis, number_of_discharges, crude_rate, easr) %>% 
  filter(age_group == "All",
         admission_type == "Emergency",
         sex == "All",
         hbr == "S92000003") %>% 
  mutate(diagnosis = recode(diagnosis, "TIAs and related syndromes" = "TIA")) %>% 
  group_by(financial_year, diagnosis) %>% 
  summarise(total_incidences_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = financial_year, y = total_incidences_easr, fill = diagnosis) +
  geom_col(position = "dodge") +
  theme_light(base_size = 15) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), 
        legend.position = "none",
        legend.text = element_text(size = 10)) +
  labs(
    x = "\nYear", 
    y = "incidence rate per 100k population",
    title = "Cerebrovascular Disease",
    subtitle = "age and sex adjusted incidence rates per year"
  ) +
   scale_fill_manual(
    values = c(
      "Cerebrovascular Disease" = "indianred",
      "Stroke" = "springgreen3",
      "Subarachnoid Haemorrhage" = "springgreen4",
      "TIA" = "darkgreen")
    )

plot_diagnosis_incidences

# save the plot:
ggsave(plot_diagnosis_incidences, filename = "plot_diagnosis_incidences.pdf", 
       path = "../outputs/", width = 19, height = 12, units = "cm")
```
Graph shows the total number of stroke incidences in Scotland over the years 
2012/13 - 2021/22 divided for diagnosis type. 
Majority of cases are diagnosed as Cerebrovascular disease, followed by Stroke. 
Small number of cases are diagnosed as TIA, followed by Subarachnoid Haemorrhage.
We see an initial increase until 2016 after incidence rates are 
more stable. Largest differences are found in Cerebrovascular disease and Stroke, while
TIA and Subarachnoid Haemorrhage are relatively stable across all years.

__What is the number of total stroke incidences split per admission type?__
```{r}
activity_hb %>% 
  select(financial_year, hbr, age_group, sex, admission_type, diagnosis, number_of_discharges, crude_rate, easr) %>% 
  filter(age_group == "All",
         admission_type != "All",
         sex == "All",
         hbr == "S92000003") %>% 
  group_by(financial_year, admission_type) %>% 
  summarise(total_incidences_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = financial_year, y = total_incidences_easr, fill = admission_type) +
  geom_col(position = "dodge")
```
Graph shows the total number of stroke incidences in Scotland over the years 
2012/13 - 2021/22 split for admission type. 
Largest number of incidences are transfered patients, followed by emergency cases. 
A small minority of admitted incidences are elective.

#### 3B. General insights Stroke mortality per hb

__What is the stroke mortality rate in Scotland per year?__
```{r}
mortality_hb %>% 
  select(year, hbr, age_group, sex, diagnosis, number_of_deaths) %>% 
  filter(age_group == "All",
         sex == "All",
         hbr == "S92000003",
         diagnosis == "Cerebrovascular Disease") %>% 
  group_by(year) %>% 
  summarise(total_deaths = sum(number_of_deaths)) %>% 
  ggplot() +
  aes(x = year, y = total_deaths) +
  geom_line(aes(group = 1), colour = "indianred", size = 2) +
  theme_light() +
  labs(
    x = "\nYear", 
    y = "Total deaths",
    title = "Total deaths for Cerebrovascular Disease per Year\n"
  ) +
  scale_y_continuous(limits = c(0, 7500))

#Saved for presentation: absolute stroke mortality rate in Scotland including gender:
mortality_absolute_gender <- mortality_hb %>% 
  select(year, hbr, age_group, sex, diagnosis, number_of_deaths) %>% 
  filter(age_group == "All",
         #sex == "All",
         hbr == "S92000003",
         diagnosis == "Cerebrovascular Disease") %>% 
  group_by(year, sex) %>% 
  summarise(total_deaths = sum(number_of_deaths)) %>% 
  ggplot() +
  aes(x = year, y = total_deaths, colour = sex) +
  geom_line(aes(group = sex), size = 2) +
  theme_light(base_size = 15) +
  labs(
    x = "\nYear", 
    y = "Total deaths",
    title = "Total deaths for Cerebrovascular Disease per Year\n"
  ) +
  scale_y_continuous(limits = c(0, 7500)) +
  scale_colour_manual(
    values = c(
      "All" = "indianred",
      "Males" = "springgreen3",
      "Females" = "springgreen4")
    )

ggsave(mortality_absolute_gender, filename = "mortality_absolute_gender.pdf", 
       path = "../outputs/", width = 19, height = 12, units = "cm")

```
Graph shows the total number of deaths due to strokes in Scotland over the year 2012-2021.
Total number of deaths has decreased over the years comparing start to end year, 
with a reduction of ~1000 cases

However this gives the absolute number of deaths, which are not standardised for 
age and sex composition differences in the population across years.
The population size changes and you want to standardise for this.

__age-sex standardised stroke deaths in Scotland per year__
```{r}
mortality_hb %>% 
  select(year, hbr, age_group, sex, diagnosis, number_of_deaths, easr, crude_rate) %>% 
  filter(age_group == "All",
         sex == "All",
         hbr == "S92000003") %>% 
  group_by(year) %>% 
  summarise(total_deaths_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = as.factor(year), y = total_deaths_easr) +
  geom_line(aes(group = 1), colour = "indianred", size = 2) +
  theme_light() +
  labs(
    x = "\nYear", 
    y = "sex-age adjusted death rate per 100,000 population",
    title = "Cerebrovascular Disease Standardised Death rate per Year",
    subtitle = "age and sex adjusted death rates per year"
  ) +
  scale_y_continuous(limits = c(0, 200))
```


__What is the number of deaths split per diagnosis type?__
```{r}
mortality_hb %>% 
  select(year, hbr, age_group, sex, diagnosis, number_of_deaths, easr) %>% 
  filter(age_group == "All",
         sex == "All",
         hbr == "S92000003") %>% 
  group_by(year, diagnosis) %>% 
  summarise(total_deaths_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = year, y = total_deaths_easr, fill = diagnosis) +
  geom_col(position = "dodge")
```
Graph shows the total number of deaths due to strokes in Scotland over the year 
2012-2021 split by diagnosis type. 
Majority of deaths are diagnosed as Cerebrovascular disease, followed by stroke. 
Only a minority is caused by Subarachnoid haemorrhage.

#### 3C. Stroke mortality per SIMD

Visualise the number of deaths (use age-sex standardised rate per 100k population)
```{r}
head(mortality_simd)

plot_cd_mortality_simd <- mortality_simd %>% 
  mutate(quintile = recode(quintile,
                           "1" = "Q1",
                           "2" = "Q2",
                           "3" = "Q3",
                           "4" = "Q4",
                           "5" = "Q5")) %>% 
  ggplot() +
  aes(x = as.factor(year), y = easr_rate, group = quintile, colour = quintile) +
  geom_line(size = 1) +
  scale_colour_manual(
    values = c(
      "Q1" = "indianred",
      "Q2" = "darkorange",
      "Q3" = "gold",
      "Q4" = "skyblue1",
      "Q5" = "navyblue"
    )
  ) +
  theme_light(base_size = 15) +
  labs(
    x = "\nYear",
    y = "mortality rate per 100,000 population\n",
    title = "Cerebrovascular Disease",
    subtitle = "Age-Sex standardised mortality rates by SIMD quintile and year",
    colour = "SIMD quintile")

plot_cd_mortality_simd

# save the plot:
ggsave(plot_cd_mortality_simd, filename = "cd_mortality_simd_2.pdf", 
       path = "../outputs/", width = 19, height = 12, units = "cm")
  
  
```
Graph shows the age-sex standardised mortality rates (per 100,000 people) per SIMD quintile.
Indicating that mortality rate for CD is higher in areas of higher deprivation.

#### 4A. Stroke incidence rate - demographics
Visualise the number of incidences (use age-sex standardised rate per 100k population)
split per age-group and sex

__sex__
```{r}
plot_cd_incidence_sex <- activity_hb %>% 
  select(financial_year, hbr, age_group, sex, admission_type, diagnosis, number_of_discharges, crude_rate, easr) %>% 
  filter(age_group == "All",
         admission_type == "All",
         sex == "Males" | sex == "Females",
         hbr == "S92000003",
         diagnosis == "Cerebrovascular Disease") %>% 
  group_by(financial_year, sex) %>% 
  summarise(total_incidences_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = financial_year, y = total_incidences_easr) +
  geom_line(aes(group = sex, colour = sex), size = 2) +
  theme_light(base_size = 15) +
  labs(
    x = "\nYear", 
    y = "age-sex adjusted incidence rate per 100,000 population\n",
    title = "Cerebrovascular Disease Standardised Incidences per Sex",
    subtitle = "age and sex adjusted incidence rates per year (ESP-2013)"
  ) +
  scale_y_continuous(limits = c(0, 1500)) +
  scale_colour_manual(
    values = c(
      "Males" = "springgreen3",
      "Females" = "springgreen4")
    )

plot_cd_incidence_sex

# save the plot:
ggsave(plot_cd_incidence_sex, filename = "cd_incidence_sex.pdf", 
       path = "../outputs/", width = 19, height = 12, units = "cm")
```

__age_group__
```{r}
activity_hb %>% 
  select(financial_year, hbr, age_group, sex, admission_type, diagnosis, number_of_discharges, crude_rate, easr) %>% 
  filter(age_group == "0-44 years" | age_group == "45-64 years" | 
           age_group == "65-74 years" | age_group == "75plus years",
         admission_type == "All",
         sex == "All",
         hbr == "S92000003") %>% 
  group_by(financial_year, age_group) %>% 
  summarise(total_incidences_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = financial_year, y = total_incidences_easr) +
  geom_line(aes(group = age_group, colour = age_group), size = 1.5) +
  theme_light() +
  labs(
    x = "\nYear", 
    y = "age-sex adjusted incidence rate per 100,000 population",
    title = "Cerebrovascular Disease Standardised Incidences per Age-group",
    subtitle = "age and sex adjusted incidence rates per year (ESP-2013)",
    colour = "Age group"
  ) +
  scale_y_continuous(limits = c(0, 8000)) +
  scale_colour_manual(
    values = c(
      "0-44 years" = "palegreen4",
      "45-64 years" = "palegreen3",
      "65-74 years" = "palegreen1",
      "75plus years" = "limegreen")
    )
```


#### 4B. Stroke mortality rate - demographics
Visualise the number of deaths (use age-sex standardised rate per 100k population)
split per age-group and sex

__sex__
```{r}
plot_cd_mortality_sex <- mortality_hb %>% 
  select(year, hbr, age_group, sex, diagnosis, number_of_deaths, easr, crude_rate) %>% 
  filter(age_group == "All",
         sex == "Males" | sex == "Females",
         hbr == "S92000003",
         diagnosis == "Cerebrovascular Disease") %>% 
  group_by(year, sex) %>% 
  summarise(total_deaths_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = as.factor(year), y = total_deaths_easr) +
  geom_line(aes(group = sex, colour = sex), size = 2) +
  theme_light(base_size = 15) +
  labs(
    x = "\nYear", 
    y = "death rate per 100,000 population\n",
    title = "Cerebrovascular Disease Death rate per Sex",
    subtitle = "age and sex adjusted death rates per year (ESP-2013)"
  ) +
  scale_y_continuous(limits = c(0, 400)) +
  scale_colour_manual(
    values = c(
      "Females" = "palegreen4",
      "Males" = "palegreen1")
    )

plot_cd_mortality_sex

ggsave(plot_cd_mortality_sex, filename = "cd_mortality_sex_2.pdf", 
       path = "../outputs/", width = 19, height = 12, units = "cm")

```

__age_group__
```{r}
plot_cd_mortality_age <- mortality_hb %>% 
  select(year, hbr, age_group, sex, diagnosis, number_of_deaths, easr, crude_rate) %>% 
  filter(age_group == "0-44 years" | age_group == "45-64 years" | 
           age_group == "65-74 years" | age_group == "75plus years",
         sex == "All",
         hbr == "S92000003") %>% 
  group_by(year, age_group) %>% 
  summarise(total_deaths_easr = sum(easr)) %>% 
  ggplot() +
  aes(x = as.factor(year), y = total_deaths_easr) +
  geom_line(aes(group = age_group, colour = age_group), size = 1.5) +
  theme_light() +
  labs(
    x = "\nYear", 
    y = "age-sex adjusted death rate per 100,000 population\n",
    title = "Cerebrovascular Disease Standardised Death rate per Age-group",
    subtitle = "age and sex adjusted death rates per year (ESP-2013)",
    colour = "Age group"
  ) +
  scale_y_continuous(limits = c(0, 1500)) +
  scale_colour_manual(
    values = c(
      "0-44 years" = "palegreen4",
      "45-64 years" = "palegreen3",
      "65-74 years" = "palegreen1",
      "75plus years" = "limegreen")
    )

ggsave(plot_cd_mortality_age, filename = "cd_mortality_age.pdf", 
       path = "../outputs/", width = 19, height = 12, units = "cm")

```


