---
title: "R Notebook"
output: html_notebook
---

# PHS stroke statistics spatial analysis

This notebook contains steps and analyses to perform spatial analysis of stroke statistics
across the healthboards of scotland

Datafiles, NHS health boards, ESRI Shapefile Download:
https://spatialdata.gov.scot/geonetwork/srv/eng/catalog.search#/metadata/f12c3826-4b4b-40e6-bf4f-77b9ed01dc14

saved here: notebooks/SG_NHS_HealthBoards_2019/

```{r}
library(sf)
library(tidyverse)
```

```{r}
# read in data about scotland, healthboards using the st_read function

scotland_healthboards_2019 <- st_read(
  dsn = "SG_NHS_HealthBoards_2019/",
  layer = "SG_NHS_HealthBoards_2019"
)
```

```{r}
#check class and look at the data
class(scotland_healthboards_2019)

head(scotland_healthboards_2019, 3)
```

```{r}
attr(scotland_healthboards_2019, "sf_column")
```

```{r}
# plotting some geometries
# plot(scotland_healthboards_2019)

# that works!
```

## Let's join the stroke info data together with shape data, and lets try to plot this.

## spatial visualisations in ggplot

```{r}
head(activity_popest_hb)
```


```{r}
# I've not managed to join the full df with all stroke statistics data
# but instead you will need to do filtering to get the right number per health board
# first, before visualisation.

activity_popest_hb <- read_csv("../clean_data/activity_popest_hb.csv")

# like this:
test_14 <-  activity_popest_hb %>% 
  filter(admission_type == "All",
         diagnosis == "Cerebrovascular Disease",
         year == 2012,
         sex == "All",
         age_group == "45-64 years",
         hbr != "S92000003")

scotland_healthboards_2019 %>% 
  left_join(test_14, by = c("HBCode" = "hbr")) %>% 
  ggplot() +
  aes(fill = crude_rate) +
  geom_sf() +
  scale_fill_continuous(labels = scales::label_comma())
```

## Spatial analysis with leaflet

```{r}
library(leaflet)
library(jsonlite)
library(rgdal)

scotland_healthboards_2019 <- st_read(
  dsn = "SG_NHS_HealthBoards_2019/",
  layer = "SG_NHS_HealthBoards_2019"
)

# add incidence rate to shape file:
join_incidence_shape <- scotland_healthboards_2019 %>% 
  left_join(test_14, by = c("HBCode" = "hbr")) %>% 
  mutate(crude_rate = round(crude_rate))

head(join_incidence_shape)

#transform for leaflet
join_incidence_shape_leaflet <- st_transform(join_incidence_shape, '+proj=longlat +datum=WGS84')

pal <- colorFactor(
  palette = "viridis",
  domain = join_incidence_shape_leaflet$crude_rate)
  
# sets the leaflet background map at 100% width
nm <- leaflet(width = "100%")

#add layers to the map
leaflet_map <- leaflet() %>% 
  addTiles() %>% 
  #adds the datazone shapes based on the shapefile
  addPolygons(data = join_incidence_shape_leaflet,
              #defines how much to simplify the polygon line on each zoom level
              smoothFactor = 0.3,
              #makes shapes slightly see through
              fillOpacity = 0.6,
              #Fill shapes with the defines palette above based on crude_rate
              fillColor = ~pal(crude_rate),
              # sets pop up labels for each healthboard shape
              label = ~paste0("Healthboard: ", HBCode,
                              "<br>",
                              "Crude_rate: ", crude_rate) %>% 
                #the label has to be piped into html to render html in the labels
                lapply(htmltools::HTML),
              weight = 1,
              #highlights the shape you are hoovering over and brings it to the front
              highlightOptions = highlightOptions(color = "white",
                                                  weight = 2,
                                                  bringToFront = TRUE)) %>% 
  #set the viewing pane for ther map
 # setView(-2.9, 55.9, zoom = 6) %>% 
  #adds legend for the palette
  addLegend(
    #shows the palette
    pal = pal,
    #shows the values associated with the palette
    values = join_incidence_shape_leaflet$crude_rate,
    opacity = 0.5,
    title = "Stroke incidences per 100k population")

# saving a Leaflet Map as a html widget in R
library(htmlwidgets)

saveWidget(leaflet_map, file = "../outputs/leaflet_map.html")

```









