---
title: "R Notebook"
output: html_notebook
---

# PHS Scottish Stroke Statistics

This notebook contains steps to perform a cluster analysis to cluster/group 
NHS health boards based on stroke incidence and mortality rates.

This is helpful as it will group health boards based on closest similarity.
The outcome can be used by the health boards to seek collaboration opportunities
across boards and tackle similar problems together.

```{r}
library(tidyverse)
library(cluster)
library(corrplot)
library(dendextend)
library(factoextra)
```

Since Cluster analysis (either hierarchical clustering or k-means clustering) do not 
allow repeteated row-names or identities. This analysis is not suitable for the 
dataset, as data is aggregated and occuring many times.

As alternative we are simplifying our methods, and create a single dendrogram based on 
incidence rate of all ages and sexes taken together. for more recent year.

Load the data
```{r}
activity_hb <- read_csv("../clean_data/stroke_activity_healthboard.csv")

head(activity_hb)

#prepare datafile, all groups cumulated for most recent year
activity_all_recent <- activity_hb %>% 
  filter(admission_type == "All",
         diagnosis == "Cerebrovascular Disease",
         sex == "All",
         financial_year == "2021/22",
         age_group == "All",
         hbr != "S92000003") %>% 
  mutate(hbr = recode(hbr, "S08000015" = "Ayrshire and Arran",
              "S08000016" = "Borders",
              "S08000017" = "Dumfries and Galloway",
              "S08000019" = "Forth Valley",
              "S08000020" = "Grampian",
              "S08000022" = "Highland",
              "S08000024" = "Lothian",
              "S08000025" = "Orkney",
              "S08000026" = "Shetland",
              "S08000028" = "Western Isles",
              "S08000029" = "Fife",
              "S08000030" = "Tayside",
              "S08000031" = "Greater Glasgow and Clyde",
              "S08000032" = "Lanarkshire"))

#select data required for the clustering
activity_cluster_data <- activity_all_recent %>% 
  select(hbr, number_of_discharges, crude_rate, easr) %>% 
  column_to_rownames("hbr")

head(activity_cluster_data)

#check for missing values
activity_cluster_data %>% 
  summarise(across(everything(), .fns = ~sum(is.na(.x))))

# Standardise data by scaling
activity_cluster_data_scale <- activity_cluster_data %>% 
  mutate(across(where(is.numeric), scale))

activity_cluster_data_scale
```

```{r}
# calculating distances
distance_matrix <- activity_cluster_data_scale %>% 
  dist(method = "euclidian")

# visualising the dissimilarity matrix showing distances
fviz_dist(distance_matrix)
```

```{r}
# build dendrogram
clusters <- distance_matrix %>% 
  hclust(method = "complete")

clusters %>% 
  plot(cex = 0.5, hang = -5)
```

Now do exactly the same but with mortality rates, and compare with dendogram above

```{r}
head(mortality_hb)

```




```{r}
mortality_hb <- read_csv("../clean_data/stroke_mortality_healthboard.csv")

#prepare datafile, all groups cumulated for most recent year
mortality_all_recent <- mortality_hb %>% 
   filter(diagnosis == "Cerebrovascular Disease",
         sex == "All",
         year == "2021",
         age_group == "All",
         hbr != "S92000003") %>% 
  mutate(hbr = recode(hbr, "S08000015" = "Ayrshire and Arran",
              "S08000016" = "Borders",
              "S08000017" = "Dumfries and Galloway",
              "S08000019" = "Forth Valley",
              "S08000020" = "Grampian",
              "S08000022" = "Highland",
              "S08000024" = "Lothian",
              "S08000025" = "Orkney",
              "S08000026" = "Shetland",
              "S08000028" = "Western Isles",
              "S08000029" = "Fife",
              "S08000030" = "Tayside",
              "S08000031" = "Greater Glasgow and Clyde",
              "S08000032" = "Lanarkshire"))

#select data required for the clustering
mortality_cluster_data <- mortality_all_recent %>% 
  select(hbr, number_of_deaths, crude_rate, easr) %>% 
  column_to_rownames("hbr")

head(mortality_cluster_data)

#check for missing values
mortality_cluster_data %>% 
  summarise(across(everything(), .fns = ~sum(is.na(.x))))

# Standardise data by scaling
mortality_cluster_data_scale <- mortality_cluster_data %>% 
  mutate(across(where(is.numeric), scale))

mortality_cluster_data_scale
```

```{r}
# calculating distances
distance_matrix_mortality <- mortality_cluster_data_scale %>% 
  dist(method = "euclidian")

# visualising the dissimilarity matrix showing distances
fviz_dist(distance_matrix_mortality)
```

```{r}
# build dendrogram
clusters_mortality <- distance_matrix_mortality %>% 
  hclust(method = "complete")

clusters_mortality %>% 
  plot(cex = 0.5, hang = -5)
```

```{r}
# compare both dendrogram ans see if they are similar

clustering_mortality <- clusters_mortality %>% 
  as.dendrogram() %>% 
  dendextend::set("labels_cex", 0.5)

clustering_mortality

clustering_activity <- clusters %>% 
  as.dendrogram() %>% 
  dendextend::set("labels_cex" = 0.8)

clustering_activity

dend_diff(clustering_activity, clustering_mortality)
```

```{r}
# Improve plots and draw clustering/grouping

#mortality
plot(clusters_mortality, cex = 1, hang = -1)
rect.hclust(clusters_mortality, k = 4, border = 2:5)


#activity
plot(clusters, cex = 1, hang = -1)
rect.hclust(clusters, k = 3, border = 2:5)

```







